Towards practical dynamic induction control of wind farms: analysis of optimal control dynamics and mimicry of first-row behavior

### 2018-01-14 draft.tex
Update chapter -> paper

The analysis presented in the current chapter will lead to the identification of a coherent flow mechanism in the wake of optimally controlled first-row turbines, responsible for part of the power increase in the optimal control simulations. This resulted in a new derived simplified dynamic axial induction control strategy for first-row turbines producing a robust power increase in the second row, irrespective of precise local flow details. However, it will be shown that the controls responsible for the remainder of the power gains in the optimal control simulations are tuned to specific flow conditions. This complicates the analysis of downstream rows, and has thus far impeded the derivation of similar control strategies for the latter.

The outline of the chapter is as follows: first, Section~\ref{sec:analysis_controls_analysis} presents an analysis of the control and thrust force dynamics, and performs some numerical experiments to elucidate characteristics of the optimal controls. As expected from the flow field statistics in previous chapter, it will be shown that the coherent behavior of turbines situated in the first row of the wind farm stands out from their downstream counterparts. Hereafter, Section~\ref{sec:analysis_flow_vis} identifies the shedding of vortex rings from the first row based on a flow visualization. Further, a simple sinusoidal thrust control approach is presented that successfully mimics this process with a robust increase in power extraction extraction in the second row. Next, Section~\ref{sec:analysis_intermediate} shortly discusses the behavior of the intermediate rows, i.e. turbines which have both upstream and downstream neighbors, for which similar simple control strategies have not been identified thus far. In conclusion, Section~\ref{sec:analysis_summ} summarizes the main findings of this chapter.

@@

@@

@@

The current section focuses on the optimal thrust coefficients generated by the optimal control simulations in the previous chapter, and performs numerical experiments to uncover some of the characteristics of these controls and their dependence on the local flow conditions.

@@

Figure~\ref{fig:controls} illustrates the time evolution of some of  the thrust coefficients $\cthat$ in the C3t5 case. The figure illustrates that, for all rows but the last one (i.e. R12), $\cthat$ varies significantly in time, and that the amplitudes and frequencies of these variations are somewhat higher in the upstream rows of the farm. In contrast, row 12 features only minor unsteadiness at lower frequencies, and has an increased mean value of $\cthat$. This relatively steady behavior of the last row can be explained by the fact that there are no further downstream turbines that could benefit from row 12 actively influencing local flow conditions, hence the row optimizes its own power greedily. The increase in mean $\cthat$ can be explained based on Figure~\ref{fig:ct_sweep}, which illustrates the power extraction as a function of steady $\cthat$ for unwaked turbines, subject to identical turbulent inlet as in case C3t5. Although momentum theory predicts maximal power extraction for steady uniform inflow at $\cthat = 2$, the actual optimal steady value for the ADM at the current spatial resolution lies somewhat higher at $\cthat \approx 2.4$, for which power extraction is about 1.4\% higher than at $\cthat = 2$. This behavior is related to the overprediction in ADM power due to the diffuse turbine representation on typical simulation grids discussed in Section~\ref{sec:meth_adm}: the mass flow through the rotor disk at $\cthat > 2$ is slightly too high compared to momentum theory, resulting in a  shift of optimal $\cthat$ towards somewhat higher values. Although the linear fit $C_P' = a \cthat$ introduced in Appendix~\ref{ch:app_adm} eliminates the error in maximal power extraction, it does not correct the value of the optimal $\cthat$ (note that this could be achieved through a more complex relation between $C_P'$ and $\cthat$). Even though this explains the slight increase in mean $\cthat$ for greedily optimized last-row turbines, this does not undermine the overall power gains of the overinductive cases from previous chapter, which all feature power gains well over 10\%. Returning to the more complex thrust coefficients in the other rows it is worth noting that, based on the current dataset, no statistically significant correlations between thrust coefficients of different turbines could be found. Furthermore, attempts towards linking thrust coefficient dynamics to upstream flow measurements (e.g. velocities, shear or kinetic energy) through linear regression models and random forest regressors have been unsuccessful to date.

@@

Figure~\ref{fig:spectra_thrust_cthat}a,b illustrate power spectral densities of the row-averaged thrust forces and thrust coefficients respectively. The figure illustrates that the variances of both the thrust coefficients and their resulting forces are highest in row 1. Further downstream, rows 2 to 11 have very similar spectral behavior, and row 12 shows significantly lower variability. The high-frequency slopes of around $-5$ observed both for $f_T$ and $\cthat$ indicate that force variability on short timescales is caused mainly by thrust coefficient variations, whereas the slower thrust force dynamics tend more to a $-5/3$ slope, suggesting that these are governed by the unsteadiness in the turbulence instead. Note that, even though the spectra for all rows except row 12 collapse at frequencies below $0.05$ Hz, the first-row spectrum shows a small yet significant peak at $f \approx 0.02 \dots 0.03$ Hz ($f D / U_\infty \approx 0.2 \dots 0.3$) as indicated by the purple arrow, which also translates to an increased spectral content of the first row thrust force at these frequencies. It will be shown later in this chapter that variations in the thrust coefficient around this frequency can be harnessed to increase power extraction in the second row.

@@

With this in mind, the remainder of this chapter will focus on the first and most promising link of the control chain: the turbines situated in the first row of the wind farm.

@@

The observed shedding of ring vortices seems to occur at specific flow-synchronized times to exploit natural instabilities in the original vortex sheets. Therefore, the remainder of this chapter will attempt to accomplish the same effect by simple sinusoidal thrust variations.

@@

Figure~\ref{fig:sinus_baseline} illustrates the power extraction for all cases considered. Figure~\ref{fig:sinus_baseline}a illustrates the relative power gains over the reference case. From the figure it can be seen that there is a well-defined range of values for $A$ and $St$ for which wind-farm power can be increased substantially through upstream sinusoidal thrust variations, with a maximal power increase of $\approx$ 5\% at $(St^\star, A^\star) = (0.25, 1.5)$. In order to put the technical feasibility of these parameters into context, note that a Strouhal number $St = 0.25$ corresponds here to a sine wave period of $\approx$ 50 s for a turbine with diameter $D = 100$ m and a freestream velocity $U_\infty = 8.5$ m s$^{-1}$. Furthermore, although higher than the $\ctmax$ applied in the optimal control cases of the previous chapter, the resulting maximum thrust coefficient of 3.5 could e.g. be attained by an NREL 5MW rotor with a 50\% increase in chord length (see \cite{goit2015optimal}, Appendix A). Figure~\ref{fig:sinus_baseline}b illustrates normalized power extraction by row for the reference case, the best sinusoidal case, and the first four rows of the optimal control case C3t5 from previous chapter. The figure shows that the power gain in the sinusoidal cases originates mostly from the second row, and that power in the first row is decreased by approximately 5\%. In contrast, optimal control case C3t5, in which all rows are active, also increases power in rows 3 and 4, and reduces first-row power by only 1\%. 
\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{gains_turbulent_6D_6D_6D_wide_inphase2.eps}
	\caption{Power extraction of baseline sinusoidal thrust case ($S = 6D$, $z_0/H = 10^{-4}$). \emph{a) } Relative gain in mean wind-farm power extraction over reference case as a function of sine amplitude $A$ and frequency $St$. \emph{b) } Row-averaged mean power extraction for the best sinusoidal case, the optimal control case C3t5 from Chapter~\ref{ch:opt_induction}, and the reference case, normalized by first-row reference power.\label{fig:sinus_baseline} }
\end{figure}

@@

%\section{Summary}\label{sec:analysis_summ}
\conclusions \label{sec:analysis_summ}
The current chapter provided a further analysis of the thrust coefficient control characteristics for the C3t5 optimal control case from Chapter~\ref{ch:opt_induction}.

>>>>

The analysis presented in the current paper will lead to the identification of a coherent flow mechanism in the wake of optimally controlled first-row turbines, responsible for part of the power increase in the optimal control simulations. This resulted in a new derived simplified dynamic axial induction control strategy for first-row turbines producing a robust power increase in the second row, irrespective of precise local flow details. However, it will be shown that the controls responsible for the remainder of the power gains in the optimal control simulations are tuned to specific flow conditions. This complicates the analysis of downstream rows, and has thus far impeded the derivation of similar control strategies for the latter.

The outline of the paper is as follows: first, Section~\ref{sec:analysis_controls_analysis} presents an analysis of the control and thrust force dynamics, and performs some numerical experiments to elucidate characteristics of the optimal controls. It will be shown that the coherent behavior of turbines situated in the first row of the wind farm stands out from their downstream counterparts. Hereafter, Section~\ref{sec:analysis_flow_vis} identifies the shedding of vortex rings from the first row based on a flow visualization. Further, a simple sinusoidal thrust control approach is presented that successfully mimics this process with a robust increase in power extraction extraction in the second row. Next, Section~\ref{sec:analysis_intermediate} shortly discusses the behavior of the intermediate rows, i.e. turbines which have both upstream and downstream neighbors, for which similar simple control strategies have not been identified thus far. In conclusion, Section~\ref{sec:analysis_summ} summarizes the main findings of this paper.

@@

\begin{figure}
	\includegraphics[width=\textwidth]{figure1.eps}
	\caption{Schematic overview of wind-farm optimal control methodology from MM17. \emph{a)} Control block diagram with adjoint gradient-based optimization and LES flow models illustrating data flow of (optimal) controls $\bs{\varphi}^{(\bullet)}$, system state $\bs{q}$, cost functional $\J$ and its gradient $\nabla \J$. \emph{b)} Receding horizon framework subdividing time into discrete flow advancement windows of length $T_A$ with prediction horizon $T$. Each arrow represents a forward or adjoint LES. Every window consists of an optimization stage (blue and red lines) follow by a flow advancement stage with optimal controls $\bs{\varphi}^{\bullet}$ (green lines).  \label{fig:optimization_meth}}
\end{figure}

@@

${\J} \nabla\J \bs{\varphi}^{\bullet}$

@@

The current section focuses on the optimal thrust coefficients generated by the optimal control simulations in MM17, and performs numerical experiments to uncover some of the characteristics of these controls and their dependence on the local flow conditions.

@@

Figure~\ref{fig:controls} illustrates the time evolution of some of  the thrust coefficients $\cthat$ in the C3t5 case. The figure illustrates that, for all rows but the last one (i.e. R12), $\cthat$ varies significantly in time, and that the amplitudes and frequencies of these variations are somewhat higher in the upstream rows of the farm. In contrast, row 12 features only minor unsteadiness at lower frequencies, and has an increased mean value of $\cthat$. This relatively steady behavior of the last row can be explained by the fact that there are no further downstream turbines that could benefit from row 12 actively influencing local flow conditions, hence the row optimizes its own power greedily. The increase in mean $\cthat$ can be explained based on Figure~\ref{fig:ct_sweep}, which illustrates the power extraction as a function of steady $\cthat$ for unwaked turbines, subject to identical turbulent inlet as in case C3t5. Although momentum theory predicts maximal power extraction for steady uniform inflow at $\cthat = 2$, the actual optimal steady value for the ADM at the current spatial resolution lies somewhat higher at $\cthat \approx 2.4$, for which power extraction is about 1.4\% higher than at $\cthat = 2$. This behavior is related to the overprediction in ADM power due to the diffuse turbine representation on typical simulation grids discussed in Section~\ref{sec:meth_adm}: the mass flow through the rotor disk at $\cthat > 2$ is slightly too high compared to momentum theory, resulting in a  shift of optimal $\cthat$ towards somewhat higher values. Although the linear fit $C_P' = a \cthat$ introduced in Appendix~\ref{ch:app_adm} eliminates the error in maximal power extraction, it does not correct the value of the optimal $\cthat$ (note that this could be achieved through a more complex relation between $C_P'$ and $\cthat$). Even though this explains the slight increase in mean $\cthat$ for greedily optimized last-row turbines, this does not undermine the overall power gains of the optimal control cases in MM17, which all feature power gains well over 10\%. Returning to the more complex thrust coefficients in the other rows it is worth noting that, based on the current dataset, no statistically significant correlations between thrust coefficients of different turbines could be found. Furthermore, attempts towards linking thrust coefficient dynamics to upstream flow measurements (e.g. velocities, shear or kinetic energy) through linear regression models and random forest regressors have been unsuccessful to date.

@@

Figure~\ref{fig:spectra_thrust_cthat}a,b illustrate power spectral densities of the row-averaged thrust forces and thrust coefficients respectively. The figure illustrates that the variances of both the thrust coefficients and their resulting forces are highest in row 1. Further downstream, rows 2 to 11 have very similar spectral behavior, and row 12 shows significantly lower variability. The high-frequency slopes of around $-5$ observed both for $f_T$ and $\cthat$ indicate that force variability on short timescales is caused mainly by thrust coefficient variations, whereas the slower thrust force dynamics tend more to a $-5/3$ slope, suggesting that these are governed by the unsteadiness in the turbulence instead. Note that, even though the spectra for all rows except row 12 collapse at frequencies below $0.05$ Hz, the first-row spectrum shows a small yet significant peak at $f \approx 0.02 \dots 0.03$ Hz ($f D / U_\infty \approx 0.2 \dots 0.3$) as indicated by the purple arrow, which also translates to an increased spectral content of the first row thrust force at these frequencies. It will be shown later in this paper that variations in the thrust coefficient around this frequency can be harnessed to increase power extraction in the second row.

@@

With this in mind, the remainder of this paper will focus on the first and most promising link of the control chain: the turbines situated in the first row of the wind farm.

@@

The observed shedding of ring vortices seems to occur at specific flow-synchronized times to exploit natural instabilities in the original vortex sheets. Therefore, the remainder of this paper will attempt to accomplish the same effect by simple sinusoidal thrust variations.

@@

Figure~\ref{fig:sinus_baseline} illustrates the power extraction for all cases considered. Figure~\ref{fig:sinus_baseline}a illustrates the relative power gains over the reference case. From the figure it can be seen that there is a well-defined range of values for $A$ and $St$ for which wind-farm power can be increased substantially through upstream sinusoidal thrust variations, with a maximal power increase of $\approx$ 5\% at $(St^\star, A^\star) = (0.25, 1.5)$. In order to put the technical feasibility of these parameters into context, note that a Strouhal number $St = 0.25$ corresponds here to a sine wave period of $\approx$ 50 s for a turbine with diameter $D = 100$ m and a freestream velocity $U_\infty = 8.5$ m s$^{-1}$. The resulting maximum thrust coefficient of 3.5 could e.g. be attained by an NREL 5MW rotor with a 50\% increase in chord length (see \cite{goit2015optimal}, Appendix A). Figure~\ref{fig:sinus_baseline}b illustrates normalized power extraction by row for the reference case, the best sinusoidal case, and the first four rows of the optimal control case C3t5 from MM17. The figure shows that the power gain in the sinusoidal cases originates mostly from the second row, and that power in the first row is decreased by approximately 5\%. In contrast, optimal control case C3t5, in which all rows are active, also increases power in rows 3 and 4, and reduces first-row power by only 1\%. 
\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{gains_turbulent_6D_6D_6D_wide_inphase2.eps}
	\caption{Power extraction of baseline sinusoidal thrust case ($S = 6D$, $z_0/H = 10^{-4}$). \emph{a) } Relative gain in mean wind-farm power extraction over reference case as a function of sine amplitude $A$ and frequency $St$. \emph{b) } Row-averaged mean power extraction for the best sinusoidal case, the optimal control case C3t5 from Chapter~\ref{ch:opt_induction}, and the reference case, normalized by first-row reference power.\label{fig:sinus_baseline} }
\end{figure}

@@

%\section{Summary}\label{sec:analysis_summ}
\conclusions \label{sec:analysis_summ}
The current paper provided a further analysis of the thrust coefficient control characteristics for the C3t5 optimal control case from Chapter~\ref{ch:opt_induction}.
